---
title: "Landscape Nature Recovery metric"
author: "Ewan McHenry"
format: html
editor: visual
---

# Landscape-level Metric through time

```{r libraries}
# load libraries and functions - read subscript 01 ----
source("subparts of calculation\\sub00- loading libraries and functions.R")

```

```{r config}
#| output: false
#| message: false
#| 
source("subparts of calculation\\sub01 - configuration.R") # configureation file

ts.pallette <- wt_palette[c(3, 5, 6, 8, 9, 11, 12, 13, 14)]

```

```{r load-data}
load(paste0(
  func.conect.path,
  "\\analysis outputs\\",
  "all_treescapes_all_years_lcmres", constants$lcm.res,
  "\\", paste(years.considered, collapse = "-"), "_scapes_",
  paste(substr(pulling_landscapes, 1, 2), collapse = "-"),
  ".RData"
))


# add a col for realtive to 1st year at 100 for each landscape
all_landscape_summary <- 
  all_landscape_summary %>%
  group_by(name) %>%
  mutate(lcd.ECA_relative_to_first_year = (lcd.ECA / lcd.ECA[year == min(year)]) * 100,
         euclid.ECA_relative_to_first_year = (euclid.ECA / euclid.ECA[year == min(year)]) * 100) %>%
  ungroup()  
    


```
```{r function-plotting-time-serise}
plot_landscape_metric <- function(data,
                                  metric,
                                  relative = FALSE,
                                  y_label = NULL,
                                  title = NULL,
                                  reference_line = NULL,
                                  lower_mult = 0.95,
                                  upper_mult = 1.05) {
  
  # Construct column name for relative metrics
  value_col <- if (relative) {
    paste0(metric, "_relative_to_first_year")
  } else {
    metric
  }
  
  # Default labels
  if (is.null(y_label)) {
    y_label <- if (relative) {
      paste0(metric, " relative to first year (%)")
    } else {
      metric
    }
  }
  
  if (is.null(title)) {
    title <- if (relative) {
      paste0(
        "Change in ", metric,
        " relative to first year by landscape"
      )
    } else {
      paste0(metric, " by landscape")
    }
  }
  
  # Default reference line for relative metrics
  if (is.null(reference_line) && relative) {
    reference_line <- 100
  }
  
  # Get y values
  y <- data[[value_col]]
  
  # Build main plot
  p <- data |>
    ggplot(aes(x = year, y = .data[[value_col]])) +
    geom_line(
      linewidth = 0.7,
      aes(color = name)
    ) +
    geom_point(
      size = 2,
      shape = 21,
      aes(fill = name),
      color = "black"
    ) +
    scale_y_continuous(
      limits = c(
        min(lower_mult * y, na.rm = TRUE),
        max(upper_mult * y, na.rm = TRUE)
      ),
      breaks = pretty_breaks()
    ) +
    scale_color_manual(
      name = NULL,
      values = ts.pallette,
      guide = guide_legend(nrow = 3)

    ) +
    scale_fill_manual(
      name = NULL,
      values = ts.pallette,
      guide = "none"
    ) +
    labs(
      title = title,
      x = "Year",
      y = y_label
    ) +
    theme_pubr()
  
  # Add reference line last
  if (!is.null(reference_line)) {
    p <- p +
      geom_hline(
        yintercept = reference_line,
        linetype = "dashed",
        linewidth = 0.7
      )
  }
  
  # Return plot
  p
}
```
## Relative change

```{r creae_relative_cahnge}
# add a col for realtive to 1st year at 100 for each landscape
all_landscape_summary <- 
  all_landscape_summary %>%
  group_by(name) %>%
  mutate(lcd.ECA_relative_to_first_year = (lcd.ECA / lcd.ECA[year == min(year)]) * 100,
         euclid.ECA_relative_to_first_year = (euclid.ECA / euclid.ECA[year == min(year)]) * 100,
         sum.patch.ha_relative_to_first_year = (sum.patch.ha / sum.patch.ha[year == min(year)]) * 100) %>%
  ungroup()  


```

::::: panel-tabset
## LCD ECA

::: panel-tabset
## Relative ECA

```{r p1.1.1-eca_rel}
plot_landscape_metric(
  all_landscape_summary,
  metric = "lcd.ECA",
  relative = TRUE,
  y_label = "LCD ECA relative to first year (%)")



```

## Absolute ECA

```{r p1.1.2-eca_abs}
plot_landscape_metric(
  all_landscape_summary,
  metric = "lcd.ECA",
  y_label = "LCD ECA (ha)",
  lower_mult = 0.9,
  upper_mult = 1.05
)

```
:::

## Euclidean ECA

::: panel-tabset
## Relative ECA

```{r p1.2.1-eca_rel}

plot_landscape_metric(
  all_landscape_summary,
  metric = "euclid.ECA",
  relative = TRUE,
  y_label = "Euclidean ECA relative to first year (%)"
)


```

## Absolute ECA

```{r p1.2.2-eca_abs}
plot_landscape_metric(
  all_landscape_summary,
  metric = "euclid.ECA",
  y_label = "Euclidean ECA (ha)"
)

```
:::
:::::

## ECA vs euclidean exploration

I hypothesised that sometimes the change in LCD ECA wasnt the same as change in Euclidean ECA, so I wanted to explore that a bit more.

First looking for a pattern in the actual measurement

::: panel-tabset
## Relative to year 1

```{r p2.1-eca-euclid-lcd_rel}
# scatter plot of lcd.ECA vs euclid.ECA for each landscape and year. Colour by treescape as before

all_landscape_summary |> 
  ggplot(aes(x = euclid.ECA_relative_to_first_year, y = lcd.ECA_relative_to_first_year, color = name)) +
  geom_point(size = 2, aes(alpha = year))  +
  # add a 1:1 line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(
    name = NULL,
    values = ts.pallette,
    guide = guide_legend(nrow = 3)
  ) +
  scale_alpha_continuous(
  name = "Year",
  range = c(0.3, 1),
  breaks = c(
    min(all_landscape_summary$year, na.rm = TRUE),
    median(all_landscape_summary$year, na.rm = TRUE),
    max(all_landscape_summary$year, na.rm = TRUE)
  ),
  guide = guide_legend(ncol = 1)) +
labs(
    title = "LCD ECA vs Euclidean ECA by landscape",
    x = "Euclidean ECA (ha)",
    y = "LCD ECA (ha)"
  ) +
  theme_pubr()


```


## Absolute

```{r p2.1-eca-euclid-lcd_abs}

all_landscape_summary |> 
  ggplot(aes(x = euclid.ECA, y = lcd.ECA, color = name)) +
  geom_point(size = 2, aes(alpha = year))  +
  # add a 1:1 line
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(
    name = NULL,
    values = ts.pallette,
    guide = guide_legend(nrow = 3)
  ) +
  scale_alpha_continuous(
  name = "Year",
  range = c(0.3, 1),
  breaks = c(
    min(all_landscape_summary$year, na.rm = TRUE),
    median(all_landscape_summary$year, na.rm = TRUE),
    max(all_landscape_summary$year, na.rm = TRUE)
  ),
  guide = guide_legend(ncol = 1)) +
  labs(
    title = "LCD ECA vs Euclidean ECA by landscape",
    x = "Euclidean ECA (ha)",
    y = "LCD ECA (ha)"
  ) +
  theme_pubr()



```
:::

Relative - perhaps a bit confusing, important to remember everything is relative to 100 int eh first year. When ECA  decreases, it decreases more for LCD.


Absolute - Not a perfect 1:1 relationship, the ECA increases more than the Euclidean ECA. 
- Makes sense that greater distance allows for a more efficient connectivity calculation: there are more options for an optimum route.
- Highlights the importance to compare rates of change, rather than absolute values.

The relationship appears to be somewhat different

# Habitat cover

::: panel-tabset
## Relative to year 1

```{r p3.1-habitat-cover-timeserise}
plot_landscape_metric(
  all_landscape_summary,
  metric = "sum.patch.ha",
  relative = TRUE,
  y_label = "Habitat area relative to first year (%)"
)

```

## Absolute

```{r p3.2-habitat-cover-timeserise}

plot_landscape_metric(
  all_landscape_summary,
  metric = "sum.patch.ha",
  y_label = "Habitat area (ha)"
)
```
:::

The uptick in 2024 is a little bit suspect, and I worry if it is true, or related to error in the LCM.

Table of rate of annual change by treescape.
```{r rate-of-change-table}

all_landscape_summary |> 
  group_by(name) |> 
  mutate(
    lcd.ECA_change_rate = lcd.ECA/lag(lcd.ECA),
    euclid.ECA_change_rate = euclid.ECA/lag(euclid.ECA),
    sum.patch.ha_change_rate = sum.patch.ha/lag(sum.patch.ha)
  ) |> 
  #col for treescape and each year, habitat change rate
  select(name, year, sum.patch.ha_change_rate) |>
  pivot_wider(names_from = year, values_from = sum.patch.ha_change_rate) |> 
  kableExtra::kbl(caption = "Rate of change in habitat area by treescape and year")

    
```